Cognate beginnings to bilingual lexical acquisition

### 2023-06-18 manuscript/manuscript.qmd
New version of introduction and discussion, n-th attempt to run Docker

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
ftn <- "For participants exposed to Catalan and Spanish exclusively this proportion is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    sub_missing(everything(), everything(), missing_text = "--") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:7) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan))) |>
    tab_footnote(ftn, location = cells_column_labels(doe_catalan))
```

@@

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients."
#| tbl-note: "Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution; narrowest range of values that contains 95\\% of the distribution, and therefore is 95\\% likely to contain the true value, given the data. On top of each distribution, we indicate the proportion of posterior samples in the 95\\% HDI that fall into the region of practical equivalence (ROPE). This proportion indicates the probability of the true value of the coefficient being equivalent to zero. Lower values indicate that the true value is unlikely to be zero of equivalent to zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"
ftn <- "Proportion of overlap between the ROPE and the 95% HDI of the coefficient, interpreted as the probability of the true value of the coefficient being equivalent to zero."
model_summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("*p*(ROPE)")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Thin lines correspond to 50 mean predictions. Thicker lines indicate the median of those predictions. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are show on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
clrs_reorder <- c("#58508d", "#003f5c", "#122A36", "#ffa600", "#ff6361", "#bc5090")

model_epreds |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) |> 
    ggplot(aes(age, .value, colour = lv, fill = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             colour = NA) +
    annotate(geom = "text",
             x = mean(responses$age),
             y = 0,
             hjust = -0.1,
             vjust = 0,
             label = "Observed age range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linetype = "dotted") +
    geom_hline(yintercept = 0.5,
               linetype = "dotted") +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 3/4,
              alpha = 1/20) +
    stat_summary(fun = median, geom = "line", linewidth = 0.75) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         linetype = "Cognateness (phonological similarity)",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)") +
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) +
    scale_colour_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_fill_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

@@

We suggest that parallel activation increases the rate at which bilingual children encounter learning instances for cognate words, but not for non-cognate words. We hypothesised that when a bilingual child is exposed to a word-form in presence of its referent, they activate not only its corresponding lexical representation, but also the lexical representation of its translation. This cross-language activation occurs at the phonological level, and the amount of co-activation that spreads from the spoken word to its translation is proportional to the amount of phonological similarity between both word-forms. Cognates would receive more activation from their translation than non-cognates, leading children to accumulate learning instances with cognate words at a faster rate than with non-cognate words. As a result, lexical representations of cognate words would consolidate at earlier ages than those of non-cognate words.

This account is in line with accumulator models of language acquisition. Accumulator models devise word acquisition as a continuous process in which a child accumulates information for each word until they reach some theoretical threshold, from which the child is considered to have acquired a word. The age-of-acquisition of a word is determined by the age at which the a child reaches the threshold of cumulative learning instances, and the rate at which a child accumulates learning instances with a word is a function of child-level properties (e.g., ability, amount of quantitative language exposure) and word-level properties (e.g., lexical frequency) [@hidaka2013computational]. In the present study, extended this account to the bilingual case, suggesting that the cognate facilitation effect on bilingual word acquisition is the result of a faster accumulation of learning instances driven by parallel activation. Accumulator models have been successfully applied to the study of vocabulary growth and word acquisition in previous work [@mcmurray2007defusing; @hidaka2013computational; @mollica2017how], but present some limitations that should also be considered in the context of the present study [see @bergelson2020comprehension]. Overall, accumulator models are a convenient tool to unify compatible accounts of language acquisition under a coherent, formal theoretical framework [@kachergis2022standard], and present an opportunity for investigating bilingual language aquisition.

In particular, the present study addressed a critical subject in bilingualism research: do bilingual infants accumulate learning experiences in both languages independently, or does exposure to one language impact the acquisition trajectory of the other language? In the context of lexical acquisition, the former scenario predicts that every learning instance for a given word-form contributes to the consolidation of the representation of such word in the lexicon, while the consolidation of its translation remains unaffected by such experience. In the latter scenario, a learning instance to the same word-form would contribute not only to the consolidation of the representation of such word, but also, to some extent, to the consolidation of its translation. Our findings provide strong support for an account of bilingual vocabulary growth in which the experience and learning outcomes accumulated by the child in one language impact those in the other language. Such a facilitatory cross-language mechanism might be an important piece in the puzzle of bilingual language acquisition. In particular, it may shed some light on why bilingual infants do not show relevant delays in language acquisition milestones compared to their monolingual peers, while receiving a reduced quantity of speech input in each of their languages. Our results provide some insights into this issue: infants benefited more strongly from the cognateness facilitation effect when acquiring words from the language of lower exposure than in the language of higher exposure.

>>>>

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "Information about language degree of exposure (DoE) was provided by families before filling the questionnaire. In this table, a 100\\% indicates that the participant was exclusively exposed to Catalan, and never to Spanish. A 0\\% indicates that the participant was not exposed to Catalan ever, and most of the time to Spanish. A 50\\% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100\\%, 0\\%, and 50\\% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4-month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
ftn <- "For participants exposed to Catalan and Spanish exclusively this proportion is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    sub_missing(everything(), everything(), missing_text = "--") |>
    cols_label(doe_catalan = "Catalan exposure") |>
    tab_spanner("Age (months)", 2:7) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan))) |>
    tab_footnote(ftn, location = cells_column_labels(doe_catalan))
```

@@

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients."
#| tbl-note: "Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. p(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"
ftn <- "Proportion of overlap between the ROPE and the 95% HDI of the coefficient, interpreted as the probability of the true value of the coefficient being equivalent to zero."
model_summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("*p*(ROPE)")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Lines correspond to 100 posterior-predicted medians. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are shown on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
clrs_reorder <- c("#58508d",
                  "#003f5c",
                  "#122A36", 
                  "#ffa600", 
                  "#ff6361", 
                  "#bc5090")

model_epreds |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) |> 
    ggplot(aes(age, .value, 
               colour = lv, 
               fill = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             colour = NA) +
    annotate(geom = "text",
             x = mean(responses$age),
             y = 0,
             hjust = -0.1,
             vjust = 0,
             label = "Observed age range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linetype = "dotted") +
    geom_hline(yintercept = 0.5,
               linetype = "dotted") +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 2/4,
              alpha = 1/5) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         linetype = "Cognateness (phonological similarity)",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)") +
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) +
    scale_colour_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_fill_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

@@

Accumulator models devise word acquisition as a continuous process in which the child gathers information about words by accumulating learning instances with such words. When the number of cumulative learning instances for a word reaches some theoretical threshold, the child is considered to have acquired such word. The rate at which a child accumulates learning instances with a word is a function of child-level properties (e.g., ability, amount of quantitative language exposure) and word-level properties (e.g., lexical frequency) [@hidaka2013computational]. Through statistical inference, formalised accumulator models provide meaningful information about parameters of interest like the aforementioned predictors [@mollica2017how; @kachergis2022standard], and allow to generate quantitative predictions about age-of-acquisition and vocabulary growth under compating theoretical accounts [@mcmurray2007defusing; @hidaka2013computational]. Capistalising on accumulator models, we extended this account to the bilingual case, suggesting that the cognate facilitation effect on bilingual word acquisition is the result of a faster accumulation of learning instances driven by parallel activation. We suggested that parallel activation increases the rate at which bilingual children encounter learning instances for cognate words, but not for non-cognate words. We hypothesised that when a bilingual child is exposed to a word-form in presence of its referent, they activate not only its corresponding lexical representation, but also the lexical representation of its translation. This cross-language activation occurs at the phonological level, and the amount of co-activation that spreads from the spoken word to its translation is proportional to the amount of phonological similarity between both word-forms. Cognates would receive more activation from their translation than non-cognates, leading children to accumulate learning instances with cognate words at a faster rate than with non-cognate words. As a result, lexical representations of cognate words would consolidate at earlier ages than those of non-cognate words.

These predictions address a critical subject in bilingualism research: do bilingual infants accumulate learning experiences in both languages independently, or does exposure to one language impact the acquisition trajectory of the other language? In the context of lexical acquisition, the former scenario predicts that every learning instance for a given word-form contributes to the consolidation of the representation of such word in the lexicon, while the consolidation of its translation remains unaffected by such experience. In the latter scenario, a learning instance to the same word-form would contribute not only to the consolidation of the representation of such word, but also, to some extent, to the consolidation of its translation. Our findings provide strong support for an account of bilingual vocabulary growth in which the experience and learning outcomes accumulated by the child in one language impact those in the other language. Such a facilitatory cross-language mechanism might be an important piece in the puzzle of bilingual language acquisition. In particular, it may shed some light on why bilingual infants do not show relevant delays in language acquisition milestones compared to their monolingual peers, while receiving a reduced quantity of speech input in each of their languages. Our results provide some insights into this issue: infants benefited more strongly from the cognateness facilitation effect when acquiring words from the language of lower exposure than in the language of higher exposure.
